---
title: "Advanced data.table operations"  
description: |
  Things get querysome and querysome.         
author: Erika Duan
date: 02-16-2021
preview: finalplot.png 
categories: 
  - data cleaning  
  - data.table  
  - dplyr
  - Pandas
  - R  
  - Python  
output:
  distill::distill_article: 
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = 'hide') 
knitr::knit_engines$set(python = reticulate::eng_python)  
options(scipen=999)
```


# Introduction    

This post is a continuation of [my previous comparison](https://erikaduan.github.io/posts/2021-01-30-data-table-part-1/) of R `data.table` versus `dplyr` versus Python `Pandas` operations.    

**Note:** The code used to create this dataset can be accessed from my github repository [here](https://github.com/erikaduan/R-tips/blob/master/03_blog_posts/2020-04-07_data-table-versus-dplyr/2020-04-07_data-table-versus-dplyr.md).    

```{r, message = FALSE, warning = FALSE}
#-----load required packages-----  
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here,
               ids, # generate random ids
               tidyverse,
               data.table,
               microbenchmark,
               reticulate)

#-----set up the Python reticulate engine-----  
conda_list() # list available conda environments
use_condaenv("r-reticulate")
```

```{r, echo = FALSE}
#-----set plugin to knit Python code chunks-----
py_run_string("import os as os")
py_run_string("os.environ['QT_QPA_PLATFORM_PLUGIN_PATH'] = 'C:/Users/user/Anaconda3/envs/r-reticulate/Library/plugins/platforms'")  
```

```{r, echo = FALSE}
#-----create a function to generate random dates-----  
create_start_dates <- function(start_date, end_date, n) {
  if(!is.character(start_date) | !is.character(end_date)) {
    stop("Error: start_date and end_date should be written as a string in the format of YYYY-mm-dd")
  }
  
  start_date <- as.Date(start_date, format = "%Y-%m-%d")
  end_date <- as.Date(end_date, format = "%Y-%m-%d")  
  
  if(end_date < start_date) {
    stop("Error: start_date should be earlier than end_date")
  }
  
  set.seed(111)
  new_date <- sample(seq(start_date, end_date, by = "day"),
                     n,
                     replace = T)
  return(new_date)
}  
```

We will be using the same dataset describing how students are engaging with online courses:     

+ Each student has a unique ID.  
+ There are 5 different online platforms (labelled platforms A, B, C, D and E).   
+ Students have the option of taking different courses within the same platform or switching to a different platform.     
+ Start dates are recorded when the student starts the first course in a new platform.     
+ End dates are also recorded when the student exits a platform.     

```{r, echo = FALSE}
#-----create a test dataset of 500000 course enrollments-----
# create 80000 unique student_ids and sample with replacement 500000 times  

set.seed(111)
students <- random_id(n = 150000, bytes = 4,
                      use_openssl = F) # so set.seed() works  

student_id <- sample(students, 500000,
                     replace = T) %>%
  sort()

#-----simulate 5 platforms with different market shares----- 
online_platform <- sample(LETTERS[1:5], 500000,
                          replace = T,
                          prob = c(0.35, 0.05, 0.1, 0.3, 0.2)) 

#-----create 20 unique course_names-----
courses <- c("R_beginner",
             "R_intermediate",
             "R_advanced",
             "Python_beginner",
             "Python_intermediate",
             "Python_advanced",
             "machine_learning",
             "linear_algebra",
             "statistics",
             "UX_design",
             "website_design",
             "data_mining",
             "travel_writing",
             "bread_baking",
             "pottery",
             "poetry_writing",
             "contemporary_dance",
             "carpentry",
             "metal_welding",
             "fitness_training")  

online_course <- sample(courses, 500000,
                        replace = T)  

student_courses <- tibble(index = seq(1, 500000, 1), # create an index to join by
                          student_id,
                          online_platform,
                          online_course)

#-----create platform_start_date-----  
platform_subset <- student_courses %>%
  select(index, 
         student_id,
         online_platform) %>% 
  mutate(unique_group = str_c(student_id, online_platform, sep = "-")) %>%
  group_by(student_id) %>%
  mutate(previous_unique_group = lag(unique_group, 1)) %>%
  ungroup()

# filter to keep rows representing the first time a student commences on a new platform      

platform_subset <- platform_subset %>%
  filter(is.na(previous_unique_group) | previous_unique_group != unique_group)

platform_start_date <- create_start_dates(start_date = "2016-01-01", end_date = "2019-01-01",  
                                          n = nrow(platform_subset))

platform_subset <- platform_subset %>%
  mutate(platform_start_date = platform_start_date)

#-----create provider_end_date-----
# create random course lengths 

set.seed(111)
platform_length <- runif(nrow(platform_subset),
                         min = 9, max = 60) %>%
  floor(.)

platform_subset <- platform_subset %>%
  mutate(platform_end_date = platform_start_date + platform_length)

#-----left join online_course to provider_subset by index-----
platform_subset <- platform_subset %>%
  select(index,
         platform_start_date,
         platform_end_date)     

student_courses <- left_join(student_courses,
                             platform_subset,
                             by = "index")

student_courses <- student_courses %>%
  fill(c(platform_start_date, platform_end_date), 
       .direction = "up")  

student_courses <- student_courses %>%
  select(-index)   

#TODO create course start date and course end date 
# course start date is a random number between platform start and end date for subsequent courses on same platform
```

```{r, echo = FALSE}
#-----remove all objects except student_courses-----  
rm(list = setdiff(ls(), "student_courses"))
gc()
```

```{r, results = 'markup'}
#-----examine the first few rows of data-----  
student_courses %>%
  head(6)   

#-----convert data frame to data.table-----  
setDT(student_courses)
```


# Basic operation sequence   


## Using `dplyr` versus `data.table`    

Imagine you would like to subset all fitness training courses from platform C and E and create a column to denote that these were discounted courses. In `dplyr`, you can write this as a single sequence of steps and use the `%>%` pipe to separate each individual operation.        

```{r, results = 'markup'}
#-----filter and append a column using dplyr-----    
dply_query_1 <- student_courses %>%
  filter(online_course == "fitness_training",
         online_platform %in% c("C", "E")) %>%
  mutate(percentage_discount = 5)

#-----preview the first 4 rows-----      
dply_query_1 %>% 
  select(-contains("platform_")) %>% 
  head(4) 
```

In `data.table`, placing the filter and column transformation operations in separate or a single step produces different outputs. ^[Placing the two operations in separate steps is equivalent to the `dplyr` approach above.]    

```{r, results = 'markup'}
#-----filter and append a column using data.table-----  
# keep filter and column transformation steps separate  
dt_query_1 <- student_courses[(online_course == "fitness_training") 
                              & (online_platform %chin% c("C", "E"))] %>%
  .[, percentage_discount := 5]   

#-----preview the first 4 rows-----  
# subset columns which begin with "platform_"
date_cols <- grep("^platform_", colnames(student_courses), value = T)

dt_query_1[1:4, !..date_cols]
```

```{r}
#-----the wrong way to filter and append a column using data.table-----  
# all rows are retained and only rows which meet the selection criteria are transformed 
# student_courses is also modified in reference 
dt_query_1_wrong <- student_courses[(online_course == "fitness_training") 
                                    & (online_platform %chin% c("C", "E")), 
                                    percentage_discount := 5]

#-----preview the first 4 rows-----  
dt_query_1_wrong[1:4, !..date_cols]  
```

```{r, echo = FALSE}
#-----remove percentage_discount for downstream analyses-----  
student_courses[, percentage_discount := NULL]
```


## Using `Pandas`   

In `Pandas`, chaining is currently limited to `.method()` operations, so we need to write separate lines of code to extract our dataset of interest.          

```{python}
import pandas as pd    
pd_courses = r.student_courses  

#-----filter and append a column using Pandas-----
pd_query_1 = (pd_courses.loc[(pd_courses["online_course"] == "fitness_training") &
                             (pd_courses["online_platform"].isin(["C", "E"]))]
              .copy())
pd_query_1["percentage_discount"] = 5
```

```{python, results = 'markup'}
#-----preview the first 4 rows-----  
# subset columns which begin with "platform_"  
date_cols = pd_query_1.columns[pd_query_1.columns.str.contains("platform_")]

(pd_query_1.drop(columns = date_cols)
 .head(4)
 .reset_index(drop = True))
```


# Group by and summarise data    

A common operation is to group by variable(s) to retrieve summary data from each grouping(s).   


## Using `dplyr` versus `data.table`   

Imagine that we are interested in the total number of days that each student spent on any online platform. We might obtain this by grouping on individual student IDs and calculating the total of days spent on any platform.      

```{r}
#-----calculate platform dwell length using dplyr-----  
student_courses <- student_courses %>%
  mutate(platform_dwell_length = platform_end_date - platform_start_date,
         platform_dwell_length = as.integer(platform_dwell_length))

#-----calculate platform dwell length using data.table-----
student_courses[, platform_dwell_length := platform_end_date - platform_start_date] %>%
  .[, platform_dwell_length := as.integer(platform_dwell_length)]  
```

**Note:** In `dplyr`, we need to close our sequence of operations with the function `ungroup()`, which removes object metadata on row groupings.   

```{r}
#-----group by student_id and summarise platform dwell length using dplyr-----
dply_query_2 <- student_courses %>%
  group_by(student_id) %>%
  summarise(total_dwell = sum(platform_dwell_length),
            min_dwell = min(platform_dwell_length),
            median_dwell = median(platform_dwell_length),  
            max_dwell = max(platform_dwell_length)) %>%
  ungroup()   

#-----preview the first 4 rows-----   
dply_query_2 %>%
  head(4)
```

In `data.table`, we can choose which variable(s) to group by using either `by` or `keyby` in the `by` placeholder of `DT[i, j, by]`. The additional effect of using `keyby` is that `keyby` also orders the results and creates a secondary key for faster subsetting.       

```{r}
#-----group by student_id and summarise total platform dwell length using data.table-----
dt_query_2 <- student_courses[,
                              .(total_dwell = sum(platform_dwell_length),
                                min_dwell = min(platform_dwell_length),
                                median_dwell = median(platform_dwell_length),
                                max_dwell = max(platform_dwell_length)),
                              by = student_id]

#-----preview the first 4 rows-----  
dt_query_2[1:4]  
```

The problem with this approach is that some students switch between courses on the same platform so the platform dwell length can be duplicated if we simply group by students IDs. To identify when this scenario occurs, we need to:  

+ Concatenate the online platform and platform start date to create a new column.  
+ Identify students where `length(unique(new_col)) > n()`.    

```{r}
#-----create platform_key using dplyr----- 
student_courses <- student_courses %>%
  mutate(platform_key = str_c(online_platform, platform_start_date, sep = "-"))

#-----create platform_key using data.table-----  
student_courses[,
                platform_key := do.call(paste0, c(.SD, sep = "-")),
                .SDcols = c("online_platform", "platform_start_date")]
```

**Note:** The base R function `do.call()` constructs and executes a function call from a name or a function and a list of arguments to be passed to it.  

```{r}
#-----identify students with duplicate records in dplyr-----
dply_query_3 <- student_courses %>%
  group_by(student_id) %>%
  summarise(unique_platform_signups = n_distinct(platform_key),
            total_course_signups = n()) 

#-----return student_ids as vector and preview the first 3 elements-----
dply_query_3 %>%
  filter(total_course_signups > unique_platform_signups) %>%
  pull(student_id) %>%
  .[1:3]

#-----examine duplicate records example-----
student_courses %>%
  select(-c(platform_key)) %>%
  filter(student_id == "00040980")
```

```{r}
#-----identify students with duplicate records in data.table-----  
dt_query_3 <- student_courses[,
                              .(unique_platform_signups = length(unique(platform_key)),
                                total_course_signups = .N),
                              by = student_id] 

#-----return student_ids as vector and preview the first 3 elements-----
dt_query_3[total_course_signups > unique_platform_signups, 
           student_id] %>%
  .[1:3]

#-----examine duplicate records example-----  
student_courses[student_id == "00040980",
                !"platform_key"]  
```


## Using `Pandas`      


## Group by and `lead` or `lag` operations   


# Other resources      

+ The data analysis and visualisation in Python for ecologists Software Carpentry workshop on [indexing, slicing and subsetting DataFrames in Python](https://datacarpentry.org/python-ecology-lesson/03-index-slice-subset/index.html).   

+ A great side-by-side comparison of `data.table` versus `dplyr` functions from a [blog post by Atrebas](https://atrebas.github.io/post/2019-03-03-datatable-dplyr/).            

+ A list of advanced `data.table` operations and tricks from a [blog post by Andrew Brooks](http://brooksandrew.github.io/simpleblog/articles/advanced-data-table/).   