---
title: "Advanced data.table operations"  
description: |
  Things get querysome and querysome.         
author: Erika Duan
date: 02-16-2021
preview: finalplot.png 
categories: 
  - data cleaning  
  - data.table  
  - dplyr
  - Pandas
  - R  
  - Python  
output:
  distill::distill_article: 
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = 'hide') 
knitr::knit_engines$set(python = reticulate::eng_python)  
options(scipen=999)
```


# Introduction    

This post is a continuation of [my previous comparison](https://erikaduan.github.io/posts/2021-01-30-data-table-part-1/) of R `data.table` versus `dplyr` versus Python `Pandas` operations.    

**Note:** The code used to create this dataset can be accessed from my github repository [here](https://github.com/erikaduan/R-tips/blob/master/03_blog_posts/2020-04-07_data-table-versus-dplyr/2020-04-07_data-table-versus-dplyr.md).    

```{r, message = FALSE, warning = FALSE}
#-----load required packages-----  
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here,
               ids, # generate random ids
               tidyverse,
               data.table,
               microbenchmark,
               reticulate)

#-----set up the Python reticulate engine-----  
conda_list() # list available conda environments
use_condaenv("r-reticulate")
```

```{r, echo = FALSE}
#-----set plugin to knit Python code chunks-----
py_run_string("import os as os")
py_run_string("os.environ['QT_QPA_PLATFORM_PLUGIN_PATH'] = 'C:/Users/user/Anaconda3/envs/r-reticulate/Library/plugins/platforms'")  
```

```{r, echo = FALSE}
#-----create a function to generate random dates-----  
create_start_dates <- function(start_date, end_date, n) {
  if(!is.character(start_date) | !is.character(end_date)) {
    stop("Error: start_date and end_date should be written as a string in the format of YYYY-mm-dd")
  }
  
  start_date <- as.Date(start_date, format = "%Y-%m-%d")
  end_date <- as.Date(end_date, format = "%Y-%m-%d")  
  
  if(end_date < start_date) {
    stop("Error: start_date should be earlier than end_date")
  }
  
  set.seed(111)
  new_date <- sample(seq(start_date, end_date, by = "day"),
                     n,
                     replace = T)
  return(new_date)
}  
```

We will be using the same dataset describing how students are engaging with online courses:     

+ Each student has a unique ID.  
+ There are 5 different online platforms (labelled platforms A, B, C, D and E).   
+ Students have the option of taking different courses within the same platform or switching to a different platform.     
+ Start dates are recorded when the student starts the first course in a new platform.     
+ End dates are also recorded when the student exits a platform.     

```{r, echo = FALSE}
#-----create a test dataset of 500000 course enrollments-----
# create 80000 unique student_ids and sample with replacement 500000 times  

set.seed(111)
students <- random_id(n = 150000, bytes = 4,
                      use_openssl = F) # so set.seed() works  

student_id <- sample(students, 500000,
                     replace = T) %>%
  sort()

#-----simulate 5 platforms with different market shares----- 
online_platform <- sample(LETTERS[1:5], 500000,
                          replace = T,
                          prob = c(0.35, 0.05, 0.1, 0.3, 0.2)) 

#-----create 20 unique course_names-----
courses <- c("R_beginner",
             "R_intermediate",
             "R_advanced",
             "Python_beginner",
             "Python_intermediate",
             "Python_advanced",
             "machine_learning",
             "linear_algebra",
             "statistics",
             "UX_design",
             "website_design",
             "data_mining",
             "travel_writing",
             "bread_baking",
             "pottery",
             "poetry_writing",
             "contemporary_dance",
             "carpentry",
             "metal_welding",
             "fitness_training")  

online_course <- sample(courses, 500000,
                        replace = T)  

student_courses <- tibble(index = seq(1, 500000, 1), # create an index to join by
                          student_id,
                          online_platform,
                          online_course)

#-----create platform_start_date-----  
platform_subset <- student_courses %>%
  select(index, 
         student_id,
         online_platform) %>% 
  mutate(unique_group = str_c(student_id, online_platform, sep = "-")) %>%
  group_by(student_id) %>%
  mutate(previous_unique_group = lag(unique_group, 1)) %>%
  ungroup()

# filter to keep rows representing the first time a student commences on a new platform      

platform_subset <- platform_subset %>%
  filter(is.na(previous_unique_group) | previous_unique_group != unique_group)

platform_start_date <- create_start_dates(start_date = "2016-01-01", end_date = "2019-01-01",  
                                          n = nrow(platform_subset))

platform_subset <- platform_subset %>%
  mutate(platform_start_date = platform_start_date)

#-----create provider_end_date-----
# create random course lengths 

set.seed(111)
platform_length <- runif(nrow(platform_subset),
                         min = 9, max = 60) %>%
  floor(.)

platform_subset <- platform_subset %>%
  mutate(platform_end_date = platform_start_date + platform_length)

#-----left join online_course to provider_subset by index-----
platform_subset <- platform_subset %>%
  select(index,
         platform_start_date,
         platform_end_date)     

student_courses <- left_join(student_courses,
                             platform_subset,
                             by = "index")

student_courses <- student_courses %>%
  fill(c(platform_start_date, platform_end_date), 
       .direction = "up")  

student_courses <- student_courses %>%
  select(-index)   

#TODO create course start date and course end date 
# course start date is a random number between platform start and end date for subsequent courses on same platform
```

```{r, echo = FALSE}
#-----remove all objects except student_courses-----  
rm(list = setdiff(ls(), "student_courses"))
gc()
```

```{r, results = 'markup'}
#-----examine the first few rows of data-----  
student_courses %>%
  head(6)   

#-----convert data frame to data.table-----  
setDT(student_courses)
```


# Basic operation sequence   


## Using `dplyr` versus `data.table`    

Imagine you would like to subset all fitness training courses from platform C and E and create a column to denote that these were discounted courses. In `dplyr`, you can write this as sequence of steps and use the `%>%` pipe as a shorthand to separate each separate operation.       

```{r, results = 'markup'}
#-----filter and append a column using dplyr-----    
dply_query_1 <- student_courses %>%
  filter(online_course == "fitness_training",
         online_platform %in% c("C", "E")) %>%
  mutate(percentage_discount = 5)

#-----preview the first 4 rows-----      
dply_query_1 %>% 
  select(-contains("platform_")) %>% 
  head(4) 
```

In `data.table`, placing the filter and column transformation operations in separate or a single step produces different outputs. ^[Placing the two operations in separate steps is equivalent to the `dplyr` approach above.]    

```{r, results = 'markup'}
#-----filter and append a column using data.table-----  
# keep filter and column transformation steps separate  
dt_query_1 <- student_courses[(online_course == "fitness_training") 
                              & (online_platform %chin% c("C", "E"))] %>%
  .[, percentage_discount := 5]   

#-----preview the first 4 rows-----  
# subset columns which begin with "platform_"
date_cols <- grep("^platform_", colnames(student_courses), value = T)

dt_query_1[1:4, !..date_cols]
```

```{r}
#-----the wrong way to filter and append a column using data.table-----  
# all rows are retained and only rows which meet the selection criteria are transformed 
# student_courses is also modified in reference 
dt_query_1_wrong <- student_courses[(online_course == "fitness_training") 
                                    & (online_platform %chin% c("C", "E")), 
                                    percentage_discount := 5]

#-----preview the first 4 rows-----  
dt_query_1_wrong[1:4, !..date_cols]  
```

```{r, echo = FALSE}
#-----remove percentage_discount for downstream analyses-----  
student_courses[, percentage_discount := NULL]
```


## Using `Pandas`   

In `Pandas`, chaining is currently limited to `.method()` operations, so we need to write separate lines of code to extract our dataset of interest.          

```{python}
import pandas as pd    
pd_courses = r.student_courses  

#-----filter and append a column using Pandas-----
pd_query_1 = (pd_courses.loc[(pd_courses["online_course"] == "fitness_training") &
                             (pd_courses["online_platform"].isin(["C", "E"]))]
              .copy())
pd_query_1["percentage_discount"] = 5
```

```{python, results = 'markup'}
#-----preview the first 4 rows-----  
# subset columns which begin with "platform_"  
date_cols = pd_query_1.columns[pd_query_1.columns.str.contains("platform_")]

(pd_query_1.drop(columns = date_cols)
 .head(4)
 .reset_index(drop = True))
```


# Other resources      

+ The data analysis and visualisation in Python for ecologists Software Carpentry workshop on [indexing, slicing and subsetting DataFrames in Python](https://datacarpentry.org/python-ecology-lesson/03-index-slice-subset/index.html).   